Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2019-10-05 manuscript/manuscript.tex
changes to median

\begin{center}
  \footnotesize
  \begin{tabular}{l c c c c c c}
    Study & Species & Selection & Replicates & Pop. Size & Gens. & Timepoints \\ 
  \hline
    \textcite{Kelly2019-dc}  & \emph{D. simulans} & lab & 3 & $\sim$1100 & 14 & 2 \\ \hline
    \textcite{Barghi2019-qy} & \emph{D. simulans} & lab & 10 & $\sim$1000 & 60 & 7 \\ \hline
    \textcite{Castro2019-uk} & \emph{M. musculus} & \begin{tabular}{c}tibiae length\\control\end{tabular} & \begin{tabular}{c}2\\1\end{tabular} &\begin{tabular}{c}32\\28\end{tabular}  & 20 & 2 \\
 \hline
\end{tabular}
\end{center}

We first analyzed \textcite{Barghi2019-qy}, an evolve-and-resequence study with
ten replicate populations exposed to a high temperature lab environment and
evolved for 60 generations, and sequenced every ten generations. Using the
seven timepoints and ten replicate populations, we estimated the genome-wide $6
\times 6$ temporal covariance matrix $\mathbf{Q}$ for each of the ten
replicates. Each row of these matrices represent the temporal covariance
$\cov(\Delta_{_{10}} p_s, \Delta_{_{10}} p_t)$, between the allele frequency
change (in ten-generation intervals, denoted $\Delta_{_{10}} p_t$) in some
initial reference generation $s$ (the row of the matrix), and some later
timepoint $t$ (the column of the matrix). We corrected these matrices for
biases created due to sampling noise, and normalize the entries for
heterozygosity (see Supplementary Material Sections
\ref{supp:ind-depth-var-corr} and \ref{supp:matrix-correction} for details on
the bias correction). The covariances are expected to be zero when only drift
is acting in a closed population, as only heritable variation for fitness can
create covariance between allele frequency changes \parencite{Buffalo2019-io}.
Averaging across the ten replicate temporal covariances matrices, we find
temporal covariances that are statistically significant (95\% block bootstraps
CIs do not contain zero), consistent with linked selection perturbing
genome-wide allele frequency changes over very short time periods. The
covariances between all adjacent time intervals are positive and then decay
towards zero as we look at more distant time intervals, as expected when
directional selection affects linked variants' frequency trajectories until
ultimately linkage disequilibrium and the additive genetic variance for fitness
associated with neutral alleles decays \parencite{Buffalo2019-io}. The temporal
covariances per replicate are noisier but this general pattern holds; see
Supplementary Figures \ref{suppfig:barghi-cov-panels}.
\textcite{Barghi2019-qy}'s design means that the covariances we see in adjacent
time intervals are on average ten generations apart, and given the temporal
decay in covariance we see, the covariances on shorter time-scales (e.g. if
adjacent generations had been sequenced) may well be higher yet (see
Supplementary Material Section \ref{supp:barghi-covs} for more details).

\vb{One concern is that these covariances reflect a few large-effect loci
  rather than selection on a polygenic trait. Since our covariances are
  essentially averages over loci, the covariance estimate can be strongly
  affected by a single outlier region. To test whether large outlier regions
  drive the genome-wide signal we see in the \textcite{Barghi2019-qy} data, we
  calculate the covariances in 100kb windows along the genome (we refer to
  these as windowed covariances throughout) and exclude the upper and lower
  $\alpha$ tails to calculate the trimmed average windowed covariance. As
  $\alpha$ increases to 0.5, the trimmed covariance converges to the median
  windowed covariance (by the definition of the median; see Supplementary
  Figure \ref{suppfig:barghi-trimmed-mean}). In other words, this median
  windowed covariance is a measure of the genome-wide covariance robust to
  large-effect loci; we find that the median windowed covariance $\cov(\Delta
  p_0, \Delta p_{10})$ (pooling all replicates) is also significantly different
  from zero, 0.0016 (0.95\% CIS [0.0015, 0.0017]).

  XXX put a table median for all timepoints.
 }

>>>>

\begin{center}
  \footnotesize
  \begin{tabular}{l c c c c c c}
    Study & Species & Selection & Replicates & Pop. Size & Gens. & Timepoints \\ 
  \hline
    \textcite{Kelly2019-dc}  & \emph{D. simulans} & lab & 3 & $\sim$1100 & 14 & 2 \\ \hline
    \textcite{Barghi2019-qy} & \emph{D. simulans} & lab & 10 & $\sim$1000 & 60 & 7 \\ \hline
    \textcite{Castro2019-uk} & \emph{M. musculus} & \begin{tabular}{c}tibiae length\\control\end{tabular} & \begin{tabular}{c}2\\1\end{tabular} &\begin{tabular}{c}32\\28\end{tabular}  & 20 & 2 \\
 \hline
  \end{tabular}
  \caption{\gc{A summary of the main selection studies we analyzed.}}
\end{center}

We first analyzed \textcite{Barghi2019-qy}, an evolve-and-resequence study with
ten replicate populations exposed to a high temperature lab environment and
evolved for 60 generations, and sequenced every ten generations. Using the
seven timepoints and ten replicate populations, we estimated the genome-wide $6
\times 6$ temporal covariance matrix $\mathbf{Q}$ for each of the ten
replicates. Each row of these matrices represent the temporal covariance
$\cov(\Delta_{_{10}} p_s, \Delta_{_{10}} p_t)$, between the allele frequency
change (in ten-generation intervals, denoted $\Delta_{_{10}} p_t$) in some
initial reference generation $s$ (the row of the matrix), and some later
timepoint $t$ (the column of the matrix). We corrected these matrices for
biases created due to sampling noise, and normalize the entries for
heterozygosity (see Supplementary Material Sections
\ref{supp:ind-depth-var-corr} and \ref{supp:matrix-correction} for details on
the bias correction). The covariances are expected to be zero when only drift
is acting, as only heritable variation for fitness can
create covariance between allele frequency changes \gc{in a closed population} \parencite{Buffalo2019-io}.
Averaging across the ten replicate temporal covariances matrices, we find
temporal covariances that are statistically significant (95\% block bootstraps
CIs do not contain zero), consistent with linked selection perturbing
genome-wide allele frequency changes over very short time periods. The
covariances between all adjacent time intervals are positive and then decay
towards zero as we look at more distant time intervals, as expected when
directional selection affects linked variants' frequency trajectories until
ultimately linkage disequilibrium and the additive genetic variance for fitness
associated with neutral alleles decays \parencite{Buffalo2019-io}. The temporal
covariances per replicate are noisier but this general pattern holds; see
Supplementary Figures \ref{suppfig:barghi-cov-panels}.
\textcite{Barghi2019-qy}'s design means that the covariances we see in adjacent
time intervals are on average ten generations apart, and given the temporal
decay in covariance we see, the covariances on shorter time-scales (e.g. if
adjacent generations had been sequenced) may well be higher yet (see
Supplementary Material Section \ref{supp:barghi-covs} for more details).

\gc{One concern is that these covariances reflect the localized impact
  of a few large-effect loci
  rather than selection on a polygenic trait. Since our covariances are
  essentially averages over loci, the covariance estimate can be strongly
  affected by a single outlier region. To test whether large outlier regions
  drive the genome-wide signal we see in the \textcite{Barghi2019-qy} data, we
  calculate the covariances in 100kb windows along the genome (we refer to
  these as windowed covariances throughout). We take the median
  windowed covariance as a measure of the genome-wide covariance robust to
  large-effect loci; we find that the median windowed covariance $\cov(\Delta
  p_0, \Delta p_{10})$ (pooling all replicates) is also significantly different
  from zero, 0.0016 (0.95\% CIS [0.0015, 0.0017]). This median
  covariance is also limiting result of a trimmed mean that excludes the upper and lower
  $\alpha$ tails to calculate the trimmed average windowed covariance. As
  $\alpha$ increases to 0.5, the trimmed covariance converges to the median
  windowed covariance (by the definition of the median; see Supplementary
  Figure \ref{suppfig:barghi-trimmed-mean}). Thus our genomic temporal
  covariances are non-zero due to the impact of selection on many
  genomic windows.
  XXX put a table median for all timepoints. AND REWORK TEXT
 }

% \vb{One concern is that these covariances reflect a few large-effect loci
%   rather than selection on a polygenic trait. Since our covariances are
%   essentially averages over loci, the covariance estimate can be strongly
%   affected by a single outlier region. To test whether large outlier regions
%   drive the genome-wide signal we see in the \textcite{Barghi2019-qy} data, we
%   calculate the covariances in 100kb windows along the genome (we refer to
%   these as windowed covariances throughout) and exclude the upper and lower
%   $\alpha$ tails to calculate the trimmed average windowed covariance. As
%   $\alpha$ increases to 0.5, the trimmed covariance converges to the median
%   windowed covariance (by the definition of the median; see Supplementary
%   Figure \ref{suppfig:barghi-trimmed-mean}). In other words, this median
%   windowed covariance is a measure of the genome-wide covariance robust to
%   large-effect loci; we find that the median windowed covariance $\cov(\Delta
%   p_0, \Delta p_{10})$ (pooling all replicates) is also significantly different
%   from zero, 0.0016 (0.95\% CIS [0.0015, 0.0017]).

%   XXX put a table median for all timepoints.
%  }
